---
title: "Machine Learning"
author: "Fadi Murad"
date: "July 1, 2018"
output:
    html_document:
      theme: cerulean
      toc: true
      toc_depth: 4
      toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE,comment="")
```


### 1. Ideas to Start
Before we start with modeling let us take a minute to clarify a few things about this step and imbalanced dataset issue  

####1.1. Best Performance Measure
***Is accuracy measure the best measure for imbalanced dataset classification model? Well! it's not***  

It's very common for classification model (especially with simple rule-based algorithm) to give ~90% accuracy when we have 90% of the instances fall into one class, simply because our models look at the data and cleverly decide that the best thing to do is to always predict this majority class and achieve high accuracy

*Solution!*  

Below performance measures can give more insight into the accuracy of our model

1. Sensitivity = True Positive Rate (TP/TP+FN) 
2. Specificity = True Negative Rate (TN/TN +FP)
3. Precision = (TP/TP+FP)
4. Recall = Sensitivity
5. F score (A weighted average of precision and recall) = () 2 * (Precision * Recall)/ (Precision + Recall) – It is the harmonic mean of precision and recall. It is used to compare several models side-by-side. Higher the better.
6. ROC Curves and Area Under Curve (AUC)

In this project, we will use *Area Under the Curve (AUC)* to be the main performance measure.

* Usually AUC is a very good measure to use with imbalanced data.
* We are trying to build a model that has a good prediction for both classes (0 and 1 or <50,000 and > 50,000). in other words, we don't have a preference or we don't value one class accuracy over the other and AUC can do that, unlike Sensitivity for example which focus on increasing the TP rate  


####1.2. Sampling Techniques
***Do we need sampling techniques?***  

For imbalanced data, it's recommended to change the dataset to more balanced dataset in order to use it to build our predictive model  

This change is called sampling dataset and there are a few techniques could be used, in this project, we will use over-sampling & under-sampling techniques  

We will create 2 additional sets using the mentioned methods then we will apply the classification models on the original and the new sets and evaluate the results  


#### 1.3. Algorithms
We will use the below algorithms in this project:

1. Logistic Regression: we will start with a simple classification model 
2. Random Forest: often perform well on imbalanced data 
3. naive Bayes: also a good algorithm for the imbalanced data
4. SVM

We have 2 very good packages for predictive modeling CARET and MLR, due to the limited performance of my machine I will use MLR (train function in CARET package consume a high portion of memory during the runtime and create huge size objects)

***Now after we have a clear vision let us hit the road!***  

<br />  

### 2. Machine Learning

We will create tasks for train, test, over sampling, under sampling and set our validation strategy

```{r}
library(rJava)
library(mlr)
library(lattice)
library(ggplot2)
library(FSelector)
library(ROCR)

set.seed(50)

load(file="f_tr.RData")
load(file="f_test.RData")

#create task for train and test sets
tr_Task <- makeClassifTask(data = f_tr,target = "income_level")
test_Task <- makeClassifTask(data = f_test,target = "income_level")

#Create sampling tasks in order to make the data balanced
#under-sampling 
tr_Under <- undersample(tr_Task,rate = 0.1) #keep only 10% of majority class

#over-sampling
tr_Over <- oversample(tr_Task,rate=15) #multiply minority class 15 times

#Set Validation Strategy
res_Mod <- makeResampleDesc("CV",iters=3,stratify = TRUE)

```

<br />  

Let us take a look at how important each variable to the target variable, we will use **generateFilterValuesData** function  

Education and major occupation code could provide the highest info for our model while race variable has very litter to offer here

```{r}
variable_Imp <- generateFilterValuesData(tr_Task,method = "information.gain")
plotFilterValues(variable_Imp)

```


#### 2.1 Logistic Reg Model

*Set learner for logistic regression*
```{r}
logR_learner <- makeLearner("classif.logreg",predict.type = "prob", fix.factors.prediction = TRUE)

```
Some predictive models return an error when we try to predict new dataset (test set in our case) if the new set does not have all levels of a factor.
Logistic Reg. and Naive Bayes have this issue, in order to avoid this, we can add argument *fix.factors.prediction = TRUE* to our learner  

*Cross-validation for the 3 sets (original, under, over)*

```{r cache= TRUE }

logRCV_Task<- resample(logR_learner,tr_Task,res_Mod,measures = list(auc,acc,tpr,tnr,fpr))
logRCV_Under<- resample(logR_learner,tr_Under,res_Mod,measures = list(auc,acc,tpr,tnr,fpr))
logRCV_Over<- resample(logR_learner,tr_Over,res_Mod,measures = list(auc,acc,tpr,tnr,fpr))
```

*Save performance measures*
```{r}
AgglogRCV_Task<- logRCV_Task$aggr
AgglogRCV_Under<- logRCV_Under$aggr
AgglogRCV_Over<- logRCV_Over$aggr

```

*Remove resample objects to save memory*
```{r}

remove(logRCV_Task,logRCV_Over,logRCV_Under)
```

*Show performance measures*
```{r}

cbind(as.data.frame(AgglogRCV_Task),as.data.frame(AgglogRCV_Under),as.data.frame(AgglogRCV_Over))
```

For the original and oversampling task we got the below warnings:
Warning messages:  

1: glm.fit: fitted probabilities numerically 0 or 1 occurred  

2: In predict.lm(object, newdata, se.fit, scale = 1, type = ifelse(type ==  :prediction from a rank-deficient fit may be misleading  


All the three models have almost the same AUC but it seems that the original task tends to predict the majority class very good, while it has very poor result for the minority  

over and under sampling seem to have more balanced result 

<br />  

#### 2.2 Random Forest Model

*Set learner for random forest*
```{r}
randF_learner <- makeLearner("classif.h2o.randomForest",predict.type = "prob")
```

*Cross function*
```{r cache= TRUE, results='hide'}
randFCV_Task<- resample(randF_learner,tr_Task,res_Mod,measures = list(auc,acc,tpr,tnr,fpr))
randFCV_Under<- resample(randF_learner,tr_Under,res_Mod,
                         measures = list(auc,acc,tpr,tnr,fpr))
randFCV_Over<- resample(randF_learner,tr_Over,res_Mod,measures = list(auc,acc,tpr,tnr,fpr))

```


*Save performance measures*
```{r}
AggrandFCV_Task<-randFCV_Task$aggr
AggrandFCV_Under<-randFCV_Under$aggr
AggrandFCV_Over<-randFCV_Over$aggr

```

*Remove resample objects to save memory and Show performance measures in one table*

```{r}
remove(randFCV_Over,randFCV_Task,randFCV_Under)

cbind(as.data.frame(AggrandFCV_Task),as.data.frame(AggrandFCV_Under),as.data.frame(AggrandFCV_Over))

```


As in logistic reg. original task perform bad for minority class, over-sampling has very high auc 0.99 while under-sampling has a good auc with 0.94
I am gonna train both under and over tasks to compare their performance on the test set as I suspect overfitting for over-sampling

<br />  

*Train & Prediction*

```{r cache= TRUE,results='hide'}

randF_Mod_Over <- train(randF_learner, tr_Over)
randF_Pre_Over <- predict(randF_Mod_Over,test_Task)

randF_Mod_Under <- train(randF_learner, tr_Under)
randF_Pre_Under <- predict(randF_Mod_Under,test_Task)
```

*Performance measure for train models*

```{r}

randF_Over_Mes<- mlr::performance(randF_Pre_Over,list(auc,acc,tpr,tnr,fpr))
randF_Under_Mes<-mlr::performance(randF_Pre_Under,list(auc,acc,tpr,tnr,fpr))
cbind(as.data.frame(randF_Over_Mes),as.data.frame(randF_Under_Mes))
```

<br />  

Let us visualize the ROC curve(fpr vs. tpr) and Precision vs. Recall  

There is more than one way to plot ROC curves, in this project we will use **generateThreshVsPerfData** and **plotROCCurves** functions for visualization and comparisons 

```{r fig.height= 5 , fig.width=12}
randFROCR_Th <- generateThreshVsPerfData(list(Over = randF_Pre_Over, Under = randF_Pre_Under), measures = list(tpr,fpr,ppv) )

gridExtra::grid.arrange(plotROCCurves(randFROCR_Th,measures = list(fpr,tpr),diagonal = FALSE)+ggtitle("AUC") , plotROCCurves(randFROCR_Th,measures = list(tpr,ppv),diagonal = FALSE) + ggtitle("Precision vs Recall"),ncol = 2)

```

<br />  

Both models have almost same performance measures in the test set, but over-sampling have 0.99 AUC on the training set and 0.94 on the test set while under-sampling maintains his AUC performance on both sets 0.94 which indicate that under-sampling is the more reliable model  

We will consider under-sampling model the best model in the random forest in order to compare it with other models in other algorithms

<br />  

#### 2.3 Naive Bayes

*Set learner for naive bayes*
```{r}
naiveBa_learner <- makeLearner("classif.naiveBayes",predict.type = "prob", fix.factors.prediction = TRUE)
```

*cross function*

```{r cache= TRUE, results='hide'}
naiveBaCV_Task<- resample(naiveBa_learner,tr_Task,res_Mod,measures = list(auc,acc,tpr,tnr,fpr))
naiveBaCV_Under<- resample(naiveBa_learner,tr_Under,res_Mod,measures = list(auc,acc,tpr,tnr,fpr))
naiveBaCV_Over<- resample(naiveBa_learner,tr_Over,res_Mod,measures = list(auc,acc,tpr,tnr,fpr))

```

*save performance measures*
```{r}
AggnaiveBaCV_Task<- naiveBaCV_Task$aggr
AggnaiveBaCV_Under<- naiveBaCV_Under$aggr
AggnaiveBaCV_Over<- naiveBaCV_Over$aggr

```

*Remove resample objects to save memory and Show performance measures*
```{r}
remove(naiveBaCV_Task,naiveBaCV_Over,naiveBaCV_Under)

cbind(as.data.frame(AggnaiveBaCV_Task),as.data.frame(AggnaiveBaCV_Under),as.data.frame(AggnaiveBaCV_Over))

```


All 3 models have almost same AUC *0.92* let us train all of them and check their performance then  

<br />  

*Train & Prediction*
```{r cache= TRUE,results='hide'}
naiveBa_Mod_Original <- train(naiveBa_learner, tr_Task)
naiveBa_Pre_Original <- predict(naiveBa_Mod_Original,test_Task)

naiveBa_Mod_Over <- train(naiveBa_learner, tr_Over)
naiveBa_Pre_Over <- predict(naiveBa_Mod_Over,test_Task)

naiveBa_Mod_Under <- train(naiveBa_learner, tr_Under)
naiveBa_Pre_Under <- predict(naiveBa_Mod_Under,test_Task)

```

*Performance measure for train models*
```{r}
naiveBa_Original_Mes<- mlr::performance(naiveBa_Pre_Original,list(auc,acc,tpr,tnr,fpr))
naiveBa_Over_Mes<- mlr::performance(naiveBa_Pre_Over,list(auc,acc,tpr,tnr,fpr))
naiveBa_Under_Mes<-mlr::performance(naiveBa_Pre_Under,list(auc,acc,tpr,tnr,fpr))

cbind(as.data.frame(naiveBa_Original_Mes),as.data.frame(naiveBa_Over_Mes),as.data.frame(naiveBa_Under_Mes))
```

<br />  

Let us visualize the ROC curve(fpr vs. tpr) and Precision vs. Recall  

```{r fig.height= 5 , fig.width=12}
naiveBaROCR_Th <- generateThreshVsPerfData(list(Original =naiveBa_Pre_Original, Over = naiveBa_Pre_Over, Under = naiveBa_Pre_Under),
                                         measures = list(tpr,fpr,ppv) )
gridExtra::grid.arrange(plotROCCurves(naiveBaROCR_Th,measures = list(fpr,tpr),diagonal = FALSE)+ggtitle("AUC") ,
                        plotROCCurves(naiveBaROCR_Th,measures = list(tpr,ppv),diagonal = FALSE) + ggtitle("Precision vs Recall"),ncol = 2)

```

<br />  

It seems original model is slightly better than other models (especially with tpr measure). We will consider the original model as the best one on the naive bayes  

However, let us see how will SVM will perform  

<br />  


#### 2.4 SVM Model

On process :)
