---
title: "Data Exploration"
author: "Fadi Murad"
date: "June 1, 2018"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE ,message=FALSE,comment="")
```

  
This R Markdown document contains the data exploration step for this project which include loading data, prepare data and taking a look at each variable in order to understand the data   
For this project, we’ve taken the data set from UCI Machine Learning Repository, you can find zip files (train set & test set) in the folder *1-Data_Set* this files contains column names and ready to be loaded, but due to the huge sizes of this files I'v already created train and test RData files to be used in this project.    

First, we will load the train and test set and take a look, we will call the train dataset *tr* and the test set *test* and we will use it for the whole project  

```{r}

  
#load needed package for this step:
library(readr)
library(rmarkdown)
library(dplyr)
library(ggplot2)
library(tinytex)
library(gridExtra)
library(corrplot) 

#import train data set
load(file="tr.RData")
load(file="test.RData")
```
```{r eval=FALSE}
#View train set
View(tr)
str(tr)
dim(tr)

str(test)
```

  
the train data has 199,523 rows and 41 columns while the train set has 99,762 rows and 41 columns, also we can notice that most of the variables class needed to be modified, but first let us start with the target variable (income_level)

```{r}

unique(tr$income_level)
prop.table(table(tr$income_level)) #value distribution
summarise(group_by(tr,income_level),n=n()) #values count

unique(test$income_level)
prop.table(table(test$income_level)) #value distribution
summarise(group_by(test,income_level),n=n()) #values count

```


  
As this is a binary classification problem, I will encode the income_level to 0 & 1 instead of -50000 & +50000 (please note there is a difference in the format of this column between train and test set) also i will add new income_level_1 column but with logical class (TRUE & FALSE), each column will be useful for different usage in the following steps.

```{r}
#Changing income level values to 0 and 1
tr$income_level[tr$income_level==-50000] <- 0
tr$income_level[tr$income_level==+50000] <- 1

test$income_level[test$income_level=="-50000"] <- 0
test$income_level[test$income_level=="50000+."] <- 1

```
  
Now let us go back to the other variables in the data set, as we noticed before (using str function) there is a need to change the class of most of the variables according to thier nature 

```{r}
#Creating vectors contains the columns index for categorical and numeric columns 
catcol <- c(2:5,7,8:16,20:29,31:38,40)
numcol <- c(setdiff(1:40,catcol),41)

#set the class for each column
tr[numcol]<-  lapply(tr[numcol], as.numeric)
tr[catcol] <- lapply(tr[catcol], as.factor)

test[numcol]<-  lapply(test[numcol], as.numeric)
test[catcol] <- lapply(test[catcol], as.factor)

```

  
Before starting with visualizing the data and creating plots for each variables let us take a look at the categorical variables that have a huge numbers of levels and try to re-group this variables to less number of categories if possible

```{r}
#creating class_of_worker_Level column (regrouping levels) 
tr<- mutate(tr, class_of_worker_Level = ifelse(grepl('gov', tr$class_of_worker), "Government",
       ifelse(grepl('Self', tr$class_of_worker), "Self-employed",as.character(tr$class_of_worker))))

#creating Education_Level column (regrouping levels) 
tr<- mutate(tr, Education_Level = ifelse(grepl('th grade', tr$education), "the first through twelfth grades",
        ifelse(grepl('Associates degree', tr$education), "Associates degree",as.character(tr$education))))

#creating d_household_summary_Level column (regrouping levels) 
tr<- mutate(tr, d_household_summary_Level = ifelse(grepl('Child under 18', tr$d_household_summary), "Child under 18",
                                         ifelse(grepl('householder', tr$d_household_summary), "householder and relative/non-relative to householder",
                                                as.character(tr$d_household_summary))))

tr$class_of_worker_Level <- as.factor(tr$class_of_worker_Level)
tr$Education_Level <- as.factor(tr$Education_Level)
tr$d_household_summary_Level <- as.factor(tr$d_household_summary_Level)

load(file="countries.RData")
#creating continent_self,continent_father, continent_mother columns using mapping data.fram "countries", to use continent column instead of the country column
countries[1:6] <- lapply(countries, as.factor)
tr <- dplyr::left_join(tr,countries[c(1,4)], by= "country_self")
tr <- dplyr::left_join(tr,countries[c(2,5)], by= "country_father")
tr <- dplyr::left_join(tr,countries[c(3,6)], by= "country_mother")
```
  
*you can find the above **countries** data.fram in excel format in "1-Data_Set" mapping file regroup countries columns to 7 levels (according to thier continent) instead of the original 43 levels  
*  
  
Also, I will create a function that return the zero value percentage in any variable to help us to understand the importance of this variable.

```{r}
Zero_value <- function(a){
  summarise(filter(tr,a == 0),Zero_Per=n()) / summarise(tr,Zero_Count=n())
}

```


##Visualizing Variable:
**Note**: the term high/low percentage of people with income level (1) means that total of income level 1 devided by total people in the same category, then I compare this percentage to the other categories percetages and the overall income level 1 percetage *(which is 6%)* 

###Numeric Variables:

####1. Age:  
We can notice from the 1st plot the declining in the frequency after age 50, in the 2nd plot we can see that most of the people with income_level 1 fall into the age between 25 and 75 years  
Also we can notice the data has outliers at age 90


```{r}
ggplot(tr,aes(x=age,y=..density..)) +
  geom_histogram(color="white",binwidth = 1) + geom_density(col="red",size = 1)+
  theme(panel.background = element_blank()) + ggtitle("Age Varibale Density")

ggplot(tr,aes(x=age,y=..count.., fill = as.factor(income_level))) +
  geom_bar() + theme(panel.background = element_blank()) + labs(fill = "income level")+
  ggtitle("age vs. income level")

```


####2.wage_per_hour  
for this variable *94%* of the data has zero value which is very high number,
also we can notice that most of the people with income level 1 has a wage_per_hour beween *1000* and *4000*  

```{r}
ggplot(tr,aes(x=age,y= wage_per_hour)) +
  geom_jitter(aes(col=as.factor(income_level),alpha = 0.5, shape = as.factor(income_level))) + theme(panel.background = element_blank())+ labs(fill = "income level")+
  scale_y_continuous("wage per hour", breaks = seq(0,10000,1000)) +
  ggtitle("Age vs. Wage per hour per Income Level")

Zero_value(tr$wage_per_hour)

```

####3. capital_gains  
*96%* of the data in this variable has zero value  
The first plot shows the capital gain (I used Log2 for better visualization), Age, and Income_level distribution

```{r}
Zero_value(tr$capital_gains)

ggplot(tr,aes(x=capital_gains,y=age,col=as.factor(income_level),alpha=0.5)) +
  geom_point(position="jitter")+scale_x_continuous(trans='log2') +labs(fill = "income level")+
  ggtitle("Capital Gain (log2) vs. Age per Income Level")

```

As the majority of the values in this variable is zero let us plot this variable in different way (zero income and more than zero)  
Almost half of the people with capital gain tends to get +50000 income which is not the case for the people with zero capital gain 
*Note: we will use the zero and more than zero levels to plot the next few numeric variables for the same reason*

```{r}
ggplot(tr,aes(x=ifelse(capital_gains == 0,"Zero","MoreThanZero"),fill=as.factor(income_level))) + geom_bar(position="dodge")+ theme(panel.background = element_blank())+labs(fill = "income level")+
  labs(x="capital gain")+ggtitle("Capital Gain per Income Level") 



```



####4. capital_losses  

*98%* of the data has zero value
Also here, almost half of the people with capital losses tends to get +50000 income which is not the case for the people with zero capital gain 

```{r}
Zero_value(tr$capital_losses) 

ggplot(tr,aes(x=capital_losses,y=age,col=as.factor(income_level),alpha=0.5)) +
  geom_point(position="jitter")+scale_x_continuous(trans='log2') +
  theme(panel.background = element_blank())+labs(fill = "income level")+
  ggtitle("capital losses (log2) vs. age per income level")

ggplot(tr,aes(x=ifelse(capital_losses == 0,"Zero","MoreThanZero"),fill=as.factor(income_level))) +geom_bar(position="dodge")+
  theme(panel.background = element_blank())+labs(fill = "income level")+
  labs(x="capital losses")+ggtitle("capital losses per income level") 

```


####5. dividend_from_Stocks  
*89%* of the data has zero value
Almost one third of the people with dividend tends to get +50000, which is a high pecentage comparing with the the zero dividend category 


```{r}
Zero_value(tr$dividend_from_Stocks)

ggplot(tr,aes(x=dividend_from_Stocks,y=age,col=as.factor(income_level),alpha=0.5)) +
  geom_point(position="jitter")+scale_x_continuous(trans='log2') +
  theme(panel.background = element_blank())+labs(fill = "income level")+
  ggtitle("dividend from Stocks (log2) vs. age per income level")

ggplot(tr,aes(x=ifelse(dividend_from_Stocks == 0,"Zero","MoreThanZero"),fill=as.factor(income_level))) + geom_bar(position="dodge")+
  theme(panel.background = element_blank())+labs(fill = "income level")+
  labs(x="dividend from stocks")+ggtitle("dividend from Stocks per income level") 

```



####6. num_person_Worked_employer
```{r}
ggplot(tr,aes(x=num_person_Worked_employer,y=..count.., fill = as.factor(income_level))) + geom_bar(position="dodge")+ scale_x_continuous("num_person_Worked_employer", breaks = seq(0,6,1)) + theme(panel.background = element_blank())+labs(fill = "income level")

Zero_value(tr$num_person_Worked_employer) #48% zero values

```



####7. weeks_worked_in_year
For better visualization and becuase most of the values in this variable fall into 0 and 52 ,I grouped all the other values into one group (1 to 51 weeks)
People who work 52 weeks has higher probability of making +50000 income

```{r}
ggplot(tr,aes(x=ifelse(weeks_worked_in_year == 0,"0",ifelse(weeks_worked_in_year==52,52,"1 to 51 week(s)")),y=..count.., fill = as.factor(income_level))) + 
  geom_bar(position="dodge")+theme(panel.background = element_blank())+
  labs(fill = "income level")+labs(x="dividend from stocks")

```

####Correlation between numeric variables 
Below we can see a visualization for the correlation matrix between the numeric variables.  
the target variable doesn't have a good correlation with any numeric variables, the best correlation for the target variable is with weeks_worked_in_year and capital_gains (0.26 & 0.24 respectively)

```{r}
numCor<-cor(tr[numcol])
corrplot(numCor, type = "upper", order = "hclust", 
         tl.col = "black")
```

###Categorical Variables

#### 1. class_of_worker
We can see that the majority of income level 1 fall into the private category  
The self-employed category has the highest percentage of people who has income level 1 (income level 1 / total people in the same category)

```{r}
ggplot(tr,aes(x = tr$class_of_worker_Level, fill = as.factor(income_level))) +
  geom_bar(position="dodge")+
  theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+labs(fill = "income level")

```


#### 2. major_occupation_code & occupation_code
it's clear that some occupations have higher percentage of income level 1, specially: Executive admin and managerial, profissional specialty and, sales  
We can dig deeper into each major occupation and its secondary occupation codes in the 2nd plot 

```{r}
ggplot(tr,aes(x = tr$major_occupation_code, fill = as.factor(income_level))) +
  geom_bar(position="dodge")+
  theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 70,hjust = 1,size=9))+labs(fill = "income level")

ggplot(tr,aes(x = occupation_code, fill = as.factor(income_level))) + geom_bar(position="dodge")+
  theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+ 
  facet_wrap(~major_occupation_code,scales = "free",ncol = 3)+labs(fill = "income level")

```


#### 3. education
More than 50% of the people who have Doctorate or prof school degrees have income level 1.
also we can say that master and Bachelors degrees has a high percentage of people with income level 1 coparing to rest of the degrees

```{r}
ggplot(tr,aes(x =tr$Education_Level ,fill= as.factor(income_level))) + geom_bar(position="dodge")+
  theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 70,hjust = 1,size=9)) + ggtitle("education vs. income level")+labs(fill = "income level")
```


#### 4. enrolled_in_edu_inst_lastwk
No obvious trend
```{r}
ggplot(tr,aes(x = tr$enrolled_in_edu_inst_lastwk, fill = as.factor(income_level))) + 
  geom_bar(position="dodge")+
  theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+ggtitle("enrolled in edu inst lastwk vs. income level")+labs(fill = "income level")

```


#### 5. marital_status
The never-married category has low percent of the income level 1 which is understandable becuase the category contains people under 18, while married and divorced categories have a high percent of income level 1

```{r}
ggplot(tr,aes(x = tr$marital_status, fill = as.factor(income_level))) + 
  geom_bar(position="dodge")+
  theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+  ggtitle("marital status per income level")+labs(fill = "income level")

```


#### 6. major_industry_code
```{r}
ggplot(tr,aes(x = tr$major_industry_code, fill = as.factor(income_level))) +
  geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 70,hjust = 1,size=9))+
  ggtitle("major_industry_code per income level")+labs(fill = "income level")

```


#### 7- race
more than *170,000* records out of *200,000* fall into one category which makes this variable hard to include in the selected variable for our model

```{r}
ggplot(tr,aes(x = tr$race, fill = as.factor(income_level))) +
  geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+
  ggtitle("race per income level")+labs(fill = "income level")

```


#### 8- hispanic_origin
we can notice that all the hispanic origin people have income level 0, but again it's hard to include this variable in the model becuase the numbers of hispanic people is relativly small in this data set

```{r}
ggplot(tr,aes(x = tr$hispanic_origin, fill = as.factor(income_level))) +
  geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+
  ggtitle("hispanic origin per income level")+labs(fill = "income level")

```


#### 9- sex
it's obvious the men have a higher percent of people with income level 1 (almost twice as women)

```{r}
ggplot(tr,aes(x = tr$sex, fill = as.factor(income_level))) + geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+
  ggtitle("sex per income level")+labs(fill = "income level")

```



#### 10- member_of_labor_union
No obvious pattern here
```{r}
ggplot(tr,aes(x = tr$member_of_labor_union, fill = as.factor(income_level))) + geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+
  ggtitle("member of labor union per income level")+labs(fill = "income level")

```


#### 11- reason_for_unemployment

```{r}
ggplot(tr,aes(x = tr$reason_for_unemployment, fill = as.factor(income_level))) + geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+
  ggtitle("reason for unemployment per income level")+labs(fill = "income level")

```


#### 12- full_parttime_employment_stat
I think we have wrong numbers here, becuase comparing the the numbers of children and armed force category with for example the numbers in the in the major industry code variable we can notice huge differences
which make this variable unreliable for our model

```{r}
ggplot(tr,aes(x = tr$full_parttime_employment_stat, fill = as.factor(income_level))) + geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+
  ggtitle("full part time employment stat per income level")+labs(fill = "income level")

```


#### 13. tax_filer_status
No obvious pattern here
```{r}
ggplot(tr,aes(x = tr$tax_filer_status, fill = as.factor(income_level))) + geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+
  ggtitle("tax filer status per income level")+labs(fill = "income level")

```


#### 14. region_of_previous_residence
No obvious pattern here
```{r}
ggplot(tr,aes(x = tr$region_of_previous_residence, fill = as.factor(income_level))) + geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+
  ggtitle("region of previous residence per income level")+labs(fill = "income level")

```


#### 15. state_of_previous_residence
183,750 (92%) out of 200,000 records doesn't have a previous residence

#### 16. d_household_summary
it's clear that most of the people with income level 1 fall under 2 categories: Householder and Spouse of householder 
```{r}
ggplot(tr,aes(x = tr$d_household_summary_Level, fill = as.factor(income_level))) + geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 70,hjust = 1,size=9))+
  ggtitle("d_household_summary per income level")+labs(fill = "income level")

```


#### 17. migration_msa

```{r}
ggplot(tr,aes(x = tr$migration_msa, fill = as.factor(income_level))) + geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+
  ggtitle("migration msa per income level")+labs(fill = "income level")

```


#### 18. migration_reg
Almost same distribution for the income level between the categories

```{r}
ggplot(tr,aes(x = tr$migration_reg, fill = as.factor(income_level))) + geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+
  ggtitle("migration reg per income level")+labs(fill = "income level")

```


#### 19. migration_within_reg
Almost same distribution for the income level between the categories

```{r}
ggplot(tr,aes(x = tr$migration_within_reg, fill = as.factor(income_level))) + geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+
  ggtitle("migration within reg per income level")+labs(fill = "income level")

```


#### 20. live_1_year_ago
Almost same distribution for the income level between the categories, no obvious pattern

```{r}
ggplot(tr,aes(x = tr$live_1_year_ago, fill = as.factor(income_level))) + geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9)) + 
  ggtitle("live 1 year ago per income level")+labs(fill = "income level")

```


#### 21. migration_sunbelt
almost same distribution for the income level between the categories

```{r}
ggplot(tr,aes(x = tr$migration_sunbelt, fill = as.factor(income_level))) + geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+
  ggtitle("migration sunbelt per income level")+labs(fill = "income level")

```


#### 22. family_members_under_18
We can ignor this variable as we can depend on the age variable for the same pattern (all childern have income level 0)
```{r}
ggplot(tr,aes(x = tr$family_members_under_18, fill = as.factor(income_level))) + geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+
  ggtitle("family members under 18 per income level")+labs(fill = "income level")

```

#### 23. country_father
While eourp has slightly better pecentage for income level 1, central and south america has no people with income level 1

```{r}
ggplot(tr,aes(x = tr$continent_father, fill = as.factor(income_level))) + geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+
  ggtitle("continent father per income level")+labs(fill = "income level")

```

#### 24. country_mother
As same as the country father variable, eourp has slightly better pecentage for income level 1, central and south america has no people with income level 1
```{r}
ggplot(tr,aes(x = tr$continent_mother, fill = as.factor(income_level))) + geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+
  ggtitle("continent mother per income level")+labs(fill = "income level")

```

#### 25. country_self
central and south america has no people with income level 1, while the others has almost the same distribution regarding the income level 
```{r}
ggplot(tr,aes(x = tr$continent_self, fill = as.factor(income_level))) + geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+
  ggtitle("continent self per income level")+labs(fill = "income level")

```

#### 26. citizenship

```{r}
ggplot(tr,aes(x = tr$citizenship, fill = as.factor(income_level))) + geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+
  ggtitle("citizenship per income level")+labs(fill = "income level")

```

#### 27. business_or_self_employed

```{r}
ggplot(tr,aes(x = tr$business_or_self_employed, fill = as.factor(income_level))) + geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+
  ggtitle("business or self employed per income level")+labs(fill = "income level")

```

#### 28. fill_questionnaire_veteran_admin
More than 97% of the data fall into one category, we will not consider this variable for our model
```{r}
ggplot(tr,aes(x = tr$fill_questionnaire_veteran_admin, fill = as.factor(income_level))) + geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+
  ggtitle("fill questionnaire veteran admin per income level")+labs(fill = "income level")

```

#### 29. veterans_benefits

```{r}
ggplot(tr,aes(x = tr$veterans_benefits, fill = as.factor(income_level))) + geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+
  scale_y_continuous("veterans benefits", breaks = seq(0,150000,20000)) +
  ggtitle("veterans benefits per income level")+labs(fill = "income level")

```

#### 30. year
same distribution of the income level for both years

```{r}
ggplot(tr,aes(x = tr$year, fill = as.factor(income_level))) + geom_bar(position="dodge")+theme(panel.background = element_blank(),axis.text.x =element_text(angle  = 90,hjust = 1,size=9))+
  ggtitle("year per income level")+labs(fill = "income level")

```



